---
title: "Microbeam Peak-Valley Spatial Transcriptomics"
subtitle: "CosMx single-cell analysis of MBRT dose heterogeneity in 4T1 flank tumors"
author: "Jeffrey Szymanski"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 12
    fig_height: 8
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", dpi = 150,
  cache = FALSE
)

library(tidyverse)
library(knitr)
library(kableExtra)
library(DT)

DATA_DIR  <- "data"
PLOT_DIR  <- "plots"
IF_DIR    <- "../if_channel_check"

img <- function(path, caption = NULL, width = "100%") {
  if (file.exists(path)) {
    cat(sprintf('\n<figure style="text-align:center; margin: 1.5em 0;">\n'))
    cat(sprintf('<img src="%s" style="max-width:%s; border: 1px solid #ddd; border-radius: 4px; padding: 4px;" />\n', path, width))
    if (!is.null(caption)) cat(sprintf('<figcaption style="color:#666; font-size:0.9em; margin-top:0.5em;">%s</figcaption>\n', caption))
    cat('</figure>\n\n')
  } else {
    cat(sprintf('\n<p style="color:red;">Missing: %s</p>\n\n', path))
  }
}
```

# Executive Summary {.tabset}

This report presents the first spatial transcriptomic analysis of microbeam radiation therapy (MBRT) peak-valley biology using the NanoString CosMx platform. MBRT delivers dose through ~1 mm-spaced parallel beams, creating alternating high-dose **peaks** and low-dose **valleys** within the same tumor. We exploit this intra-tumoral dose gradient to ask: *what transcriptomic programs distinguish direct-beam exposure from bystander zones at single-cell resolution?*

**Key findings:**

1. **Peak-valley classification validated against gamma-H2AX IHC.** FOV-level manual annotation from adjacent-section H2AX staining, extrapolated via geometric stripe fitting (−25° angle, 1.05 mm spacing), achieves 89% FOV-level concordance (24/27 FOVs correctly labeled) and classifies ~150K cells into peak (~73K), valley (~76K), and transition zones.

2. **Transcriptomic stripe pattern detected in gene expression.** Cdkn1a (p21), the canonical DDR effector, shows periodic spatial variation aligned with beam geometry. Thirty-seven genes survive a within-block rotation null test (vs ~33 expected by chance at p < 0.05); the surviving set is dominated by MHC class I, IFN response, and stress genes, consistent with direct DNA damage biology rather than spatial architecture artifacts.

3. **Cell-type-specific dose-response gradients detected across cell types.** Dendritic cells (Becn1 rho = −0.14), pericytes (Cdkn1a rho = −0.08), and B cells (Cdkn1a rho = −0.04, improving to −0.15 with a 15-gene composite score) show detectable gradients. Macrophage subtypes show variable gradients (rho = −0.08 to −0.12), not uniformly flat as classical radioresistance would predict. All effect sizes are small and cell-type rankings are unstable at n=1.

4. **Morphological divergence emerges at day 2, not 4h.** MBRT median nuclear area progressively declines while SBRT nuclei enlarge at day 2, yielding ~23% divergence. This is consistent with — but not proof of — a qualitative difference in damage response (apoptosis/pyknosis in MBRT vs G2/M arrest or inflammatory edema in SBRT). Survivorship bias (selective loss of large, lethally irradiated cells from MBRT peaks) is an alternative explanation.

5. **MBRT shows higher spatial immune-tumor mixing than SBRT at 4h.** MBRT exhibits consistently higher immune neighbor fractions than SBRT across cell types (k=20 nearest neighbors). At 4h, this likely reflects redistribution of resident immune cells rather than de novo recruitment, which requires chemokine-mediated infiltration over longer timescales.

**Caveats:** All Mutter_01 analyses are descriptive (n=1 per condition). Formal statistical testing awaits Mutter_02 biological replicates. The 1000-gene targeted panel has known blind spots for radiation-specific pathway genes. Peak-valley bulk pathway score differences are negligible (Cohen's d < 0.01) and should be considered null results at the current sample size.

---

# Biological Context

## Microbeam Radiation Therapy

MBRT delivers radiation through an array of parallel, sub-millimeter-spaced planar beams. Unlike conventional spatially uniform irradiation (SBRT), MBRT creates alternating high-dose **peak** stripes and low-dose **valley** gaps across the tumor. In this experiment, the Mutter lab used mm-scale planar beams (minibeams) with ~1 mm center-to-center spacing. Prescribed peak dose, beam energy, and peak-to-valley dose ratio (PVDR) are pending from the treatment planning records; for mm-scale minibeam geometries, PVDRs of 5–15 are typical in the published literature. The resulting intra-tumoral dose heterogeneity is hypothesized to:

- **Peaks:** Induce direct DNA damage, ATM/ATR-mediated p53/p21 activation, and cell cycle arrest or apoptosis in directly irradiated cells. Secondary NF-κB signaling may drive chemokine secretion (e.g., Ccl8/MCP-2), recruiting myeloid cells to damage sites.
- **Valleys:** Receive substantially lower dose but may exhibit bystander effects mediated through gap junction communication (Cx43/Gja1), reactive oxygen/nitrogen species (ROS/RNS), and DAMP release (HMGB1, calreticulin). These signals can activate NF-κB and interferon pathways in non-irradiated cells adjacent to the beam.

An additional consideration is that the beam edges have a different energy spectrum than beam centers: scattered photons at minibeam edges have lower energy and potentially higher effective linear energy transfer (LET), meaning the biological damage quality — not just quantity — varies across the peak-valley transition. This could contribute to the observed p21 donut profile if beam-edge cells experience higher-RBE radiation producing more complex, repair-resistant DSBs per unit dose.

This spatial fractionation pattern creates a natural experiment within a single tumor section — peak and valley cells share the same tumor microenvironment, vasculature, and immune context but differ primarily in received dose. This is conceptually an "intra-tumoral abscopal" scenario where bystander and out-of-field effects can be studied within a single tissue section.

## Gamma-H2AX and DNA Damage

γ-H2AX (phosphorylated histone H2AX at Ser139) is the gold-standard immunohistochemical marker for DNA double-strand breaks (DSBs). Each nuclear focus corresponds to approximately one DSB. Kinetics after irradiation:

- **15–30 min:** γ-H2AX foci appear at break sites
- **30 min–1 hour:** Peak foci density (optimal window for dose quantification)
- **1–4 hours:** Foci gradually resolve as NHEJ and HR repair breaks
- **24 hours:** >90% of foci resolved; residual foci mark unrepaired complex lesions

The H2AX IHC staining used in this study was performed at **4 hours** post-irradiation — past peak signal and into the repair phase. Peaks are still visible but dimmer than at 30 min. This matters: the 4h H2AX image is a damped and repair-smeared snapshot of the dose distribution, not a crisp dose map.

## Cell Type Radiosensitivity

The established hematopoietic radiosensitivity hierarchy is:

- **B cells** — most radiosensitive (undergo apoptosis at low doses via p53-dependent pathways)
- **T cells** — moderately sensitive
- **Monocytes/DC** — intermediate
- **Macrophages** — radioresistant (post-mitotic, no replicative stress)

This hierarchy predicts that lymphoid cells should show sharper peak-valley transcriptomic contrast than myeloid cells, which is broadly consistent with our data (macrophage gradients are flatter than lymphoid/stromal gradients), though the fine-grained cell-type ranking is unstable at n=1.

---

# Experimental Design

## Samples and Conditions

```{r sample-table}
qc <- read_tsv(file.path(DATA_DIR, "qc_summary.tsv"), show_col_types = FALSE)

sample_info <- tibble(
  Condition = c("Control", "MBRT_1h", "MBRT_4h", "SBRT_4h",
                "MBRT_day2", "SBRT_day2", "MBRT_day6", "SBRT_day6",
                "Tongue_MBRT_day8", "Tongue_MBRT_day10", "Tongue_SBRT_day10"),
  Treatment = c("None", "MBRT", "MBRT", "SBRT",
                "MBRT", "SBRT", "MBRT", "SBRT",
                "MBRT", "MBRT", "SBRT"),
  Timepoint = c("0h", "1h", "4h", "4h", "48h", "48h", "144h", "144h",
                "192h", "240h", "240h"),
  Model = c("Flank", rep("Flank", 7), rep("Tongue", 3))
)

sample_info <- left_join(sample_info, qc, by = c("Condition" = "condition"))

sample_info %>%
  rename(`Cells (post-QC)` = post_filter) %>%
  kable(format = "html", caption = "Mutter_01 experimental conditions. All flank tumors are 4T1 mouse mammary carcinoma in Balb/c mice. n=1 per condition.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

**Tumor model:** 4T1 mouse mammary carcinoma — a triple-negative, claudin-low breast cancer line that co-expresses epithelial markers (Krt8, Krt18, Epcam) and mesenchymal markers (Vim, S100a4, S100a6) natively. The mesenchymal expression in tumor cells is real 4T1 biology, not contamination.

**Platform:** NanoString CosMx Spatial Molecular Imager with Mouse Universal Cell Characterization (UCC) panel (~950 genes) plus 21 custom add-on targets. Secondary analysis (QC, cell typing via insitutype with ImmGen reference atlas, pathway scores) performed by Yi Liu at the Mayo Spatial Biology Core.

## Gene Panel Coverage

The CosMx Mouse UCC panel provides broad immune and stromal coverage but has known gaps for radiation-specific genes. Pre-computed pathway scores cover:

- **Type I Interferon** (19 genes) — standard NanoString module
- **Type II Interferon** (17 genes) — standard NanoString module
- **DNA Damage Repair** (18 genes) — standard NanoString module
- **STING** — custom score from available cGAS-STING pathway genes

---

# Quality Control

## Cell Filtering

```{r qc-stats}
total_raw <- 1432028
total_qc <- sum(qc$post_filter, na.rm = TRUE)
cat(sprintf("**Raw cells:** %s | **Post-QC:** %s | **Retention:** %.0f%%\n\n",
            format(total_raw, big.mark = ","),
            format(total_qc, big.mark = ","),
            100 * total_qc / total_raw))
```

QC filters applied: `qcFlagsCell == "Pass"`, `nCount_RNA > 20`, `nFeature_RNA > 10`, `propNegative < 0.5`. Retention ranges from 44% (Tongue_MBRT_day8, late timepoint with extensive cell death) to 84% (Control).

```{r qc-violins, results='asis'}
img(file.path(PLOT_DIR, "qc_violins.png"),
    "QC metrics across conditions: transcript counts, detected features, and negative probe fraction.")
```

```{r qc-table}
qc %>%
  kable(format = "html", caption = "Per-condition QC summary",
        col.names = c("Condition", "Cells post-QC")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Clustering and Cell Type Validation

Normalization (LogNormalize, scale 10,000), PCA (20 dims), Louvain clustering (resolution 0.4), and UMAP produce 9 clusters. Cell types assigned by insitutype against the ImmGen mouse reference atlas are validated against 18/19 canonical markers present in the panel.

```{r umap-landscape, results='asis'}
img(file.path(PLOT_DIR, "umap_landscape.png"),
    "UMAP landscape: clusters, conditions, slide batch, and treatment.", width = "90%")
```

```{r celltype-dotplot, results='asis'}
img(file.path(PLOT_DIR, "celltype_marker_dotplot.png"),
    "Canonical marker validation of insitutype cell type labels.", width = "90%")
```

The uncharacterized **\"a\" population** (36% of cells) expresses epithelial markers (Epcam, Krt8, Krt18) at 1.4–2.2× background and is reclassified as `tumor_epithelial`. However, the \"a\" bucket is heterogeneous — only ~32% are bona fide epithelial cells (the remainder are endothelial, fibroblast, smooth muscle, and unassigned). For cancer-specific analyses, a stringent 4T1 filter is applied: mean(Krt8, Krt18, Epcam, Cdh1) > 0.3 AND mean(Pecam1, Cdh5, Vwf) < 0.1.

```{r spatial-celltype, results='asis'}
img(file.path(PLOT_DIR, "spatial_celltype_flank.png"),
    "Spatial cell type maps for all flank conditions.", width = "90%")
```

---

# Peak-Valley Identification

## The Stripe Detection Problem

MBRT beam geometry creates parallel dose stripes across the tumor, tilted relative to the CosMx coordinate axes. The stripes must be detected and each cell classified as **peak** (direct beam), **valley** (inter-beam gap), or **transition** before any peak-valley transcriptomic comparison can be made.

Two approaches were developed:

### Geometric Stripe Model (superseded)

Initial approach: fit a periodic stripe model to transcriptomic data alone. Optimized tilt angle (15°), spacing (~1.02 mm), and phase by maximizing Cdkn1a contrast between putative zones. **Limitation:** transcript-based DDR scores are too noisy to resolve stripes at sub-mm precision — the model fit depends on the gene used for optimization and the resulting classification is circular.

### FOV Extrapolation (current method) {#fov-extrapolation}

The validated method uses manually annotated FOVs from gamma-H2AX IHC on an adjacent serial section:

1. **Manual FOV annotation:** 14 peak FOVs and 13 valley FOVs identified by visual inspection of the H2AX stain on Block_21 (MBRT 4h).
2. **Stripe angle fitting:** k-means clustering (k=4 stripes) on peak FOV centroids projected along a grid of angles identifies the beam orientation (−25° from horizontal).
3. **Extrapolation:** All ~150K MBRT_4h cells are labeled by their perpendicular distance to the nearest fitted beam centerline.
4. **Zone assignment:** Cells within `peak_half` (~0.27 mm) of a beam center = peak; cells beyond `beam_spacing/2 - peak_half` from any center = valley; remainder = transition.

```{r fov-extrapolation, results='asis'}
img(file.path(IF_DIR, "block21_extrapolated_pv.png"),
    "Cell-level peak/valley extrapolation from FOV centroids. Red = peak, blue = valley. Dashed lines = beam centerlines.", width = "80%")
```

**FOV lists:**

- **Peak FOVs:** 224, 209, 201, 186, 172, 225, 229, 215, 175, 176, 206, 181, 168, 167
- **Valley FOVs:** 199, 213, 227, 188, 204, 218, 178, 165, 170, 231, 189, 173, 154
- **Block:** 21 | **Slide:** 20250529_214712_S4

**Stripe parameters:** θ = −25°, spacing = 1.05 mm, beam half-width = 0.27 mm. FOV-level concordance: 89% (24 of 27 manually labeled FOVs classified consistently). The 3 discordant FOVs are at stripe boundaries. Cell-level classification yields ~73K peak and ~76K valley cells, accounting for essentially all ~149K MBRT_4h cells; the zone assignment parameters produce a negligibly small transition zone.

## H2AX Validation

The stripe model zone map is visually compared against the gamma-H2AX chromogenic IHC image extracted from the lab-provided PPTX. Programmatic co-registration was attempted using H2AX intensity profiles but limited by PPTX image resolution (the PPTX screenshots are not full-resolution scanner TIFFs).

**Important caveat on 4h H2AX:** At 4h post-irradiation, simple DSBs (repaired by NHEJ in ~30 min) are already resolved, while complex/clustered DSBs (requiring HR, 4–8h) persist. The 4h H2AX image is therefore biased toward regions with the most complex, hardest-to-repair damage — primarily beam peaks with the highest dose. This may cause the H2AX-defined peak boundaries to be *narrower* than the true dose peak (beam-edge cells with lower dose have already resolved their H2AX signal), potentially misclassifying some beam-edge cells as transition or valley. Adjacent serial sections can also introduce 50–100 µm tissue displacement, which is a meaningful fraction of the 0.27 mm beam half-width.

```{r h2ax-zones, results='asis'}
img(file.path(PLOT_DIR, "h2ax_zone_map_labeled.png"),
    "Stripe model zone map with FOV labels. Yellow lines = beam centerlines.", width = "80%")
```

```{r h2ax-concordance, results='asis'}
img(file.path(PLOT_DIR, "h2ax_concordance_panel.png"),
    "H2AX concordance panel: zone classification vs gamma-H2AX image.", width = "85%")
```

## Biological Consistency: p21 Stripe Recovery

As a consistency check (not an independent validation, since both the stripe model and p21 depend on the same underlying dose distribution), **Cdkn1a (p21) expression independently recovers the beam geometry from gene expression alone.**

```{r cdkn1a-panel, results='asis'}
img(file.path(PLOT_DIR, "cdkn1a_evidence_panel.png"),
    "Cdkn1a evidence panel: spatial heatmap and y-profile show periodic p21 variation aligned with beam stripes.", width = "90%")
```

```{r p21-h2ax-overlay, results='asis'}
img(file.path(PLOT_DIR, "p21_h2ax_overlay.png"),
    "p21 spatial pattern overlaid with H2AX peak zones.", width = "80%")
```

---

# MBRT vs SBRT Comparative Kinetics

Before examining peak-valley biology, we establish that MBRT and SBRT produce globally distinct transcriptomic responses.

## Differential Expression

```{r deg-summary}
degs <- read_tsv(file.path(DATA_DIR, "deg_all_cells.tsv"), show_col_types = FALSE)
cat(sprintf("Total bulk DEG rows across all comparisons: **%s**\n\n", format(nrow(degs), big.mark = ",")))

deg_counts <- degs %>%
  filter(abs(avg_log2FC) > 0.5) %>%
  count(comparison_type, comparison, name = "n_degs_lfc05") %>%
  arrange(comparison_type, comparison)

deg_counts %>%
  kable(format = "html", caption = "DEGs with |log2FC| > 0.5 per comparison",
        col.names = c("Type", "Comparison", "DEGs")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

**Top MBRT vs SBRT findings:**

- **4h:** MBRT upregulates myeloid recruitment genes (Ccl8 log2FC 1.59, Mrc1, Cd74, Arg1)
- **Day 2:** Myeloid signal amplifies (Ccl8 log2FC 2.43, Cd163 1.74); SBRT upregulates Cxcl10 (−1.53, IFN-stimulated)
- **Day 6:** MBRT shows massive Igkc (log2FC 3.47, B/plasma cell signature); SBRT has Nkg7 (−1.31, cytotoxic)

```{r volcano, results='asis'}
img(file.path(PLOT_DIR, "volcano_mbrt_vs_sbrt.png"),
    "Volcano plots: MBRT vs SBRT at 4h, day 2, and day 6. Red = |log2FC| > 0.5.", width = "90%")
```

## Pathway Score Kinetics

```{r pathway-kinetics, results='asis'}
img(file.path(PLOT_DIR, "pathway_kinetics.png"),
    "Pathway score kinetics: DDR, IFN-I, IFN-II, STING across timepoints. MBRT (red) vs SBRT (blue).", width = "80%")
```

MBRT pathways are suppressed at early timepoints (1h, 4h) relative to control, then rebound at day 2. SBRT shows a similar pattern where measured (4h suppression, day 2 rebound). At day 2, **MBRT shows stronger activation than SBRT for DDR, Type I IFN, and Type II IFN** (e.g., Type I IFN: MBRT 0.56 vs SBRT 0.21; DDR: MBRT 0.49 vs SBRT 0.28). The exception is **STING**, where SBRT exceeds MBRT at day 2. The early suppression (1–4h) is consistent with DNA damage signaling prioritizing repair over transcription — a combination of local transcription arrest at DSB-proximal genes, mRNA degradation at damaged chromatin, and the lag between protein signaling and mRNA accumulation. The day 2 rebound represents the secondary transcriptional wave once repair is complete or cells commit to downstream fate decisions.

## Cell Type Composition Shifts

```{r composition, results='asis'}
img(file.path(PLOT_DIR, "composition_stacked.png"),
    "Cell type composition across conditions. Tumor epithelial (\"a\") fraction declines post-radiation as immune cells infiltrate.", width = "80%")
```

## Spatial Immune Neighborhoods

```{r spatial-nn, results='asis'}
img(file.path(PLOT_DIR, "spatial_nn_immune_frac.png"),
    "Fraction of k=20 nearest neighbors that are immune cells, by cell type. MBRT (red) vs SBRT (blue) at 4h.", width = "90%")
```

MBRT shows consistently higher immune neighbor fractions than SBRT at 4h across every cell type — dendritic cells: 0.69 vs 0.52; microglia: 0.72 vs 0.59; even tumor cells: 0.37 vs 0.31. At 4h, this likely reflects redistribution of pre-existing tumor-infiltrating immune cells rather than de novo recruitment (insufficient time for chemokine-mediated infiltration). Possible mechanisms include differential cell death creating spatial gaps that resident immune cells fill, or radiation-induced changes in cell adhesion/motility. The chemokine signal (Ccl8 peak enrichment) may drive subsequent myeloid recruitment at later timepoints.

---

# Peak-Valley Transcriptomic Analysis (MBRT 4h)

## Bulk Differential Expression

```{r pv-degs}
pv_degs <- read_tsv(file.path(DATA_DIR, "pv_degs_bulk.tsv"), show_col_types = FALSE)

pv_degs %>%
  head(25) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(format = "html", caption = "Top 25 peak vs valley DEGs (all cells, ranked by effect size). Positive log2FC = higher in peaks.",
        col.names = c("p-value", "log2FC", "% peak", "% valley", "adj. p", "Gene")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r pv-volcano, results='asis'}
img(file.path(PLOT_DIR, "pv_volcano_bulk.png"),
    "Peak vs valley volcano plot (all cells). Positive = higher in peaks. Effect sizes are small (max |log2FC| ~0.35) reflecting n=1 and spatial mixing.", width = "70%")
```

**Interpretation:** The top peak-enriched gene is **Ccl8** (log2FC 0.34), a monocyte chemoattractant. This is consistent with expected biology: Ccl8/MCP-2 is induced by NF-κB and STAT1, both activated downstream of ATM/ATR DNA damage signaling, and recruits myeloid cells to damage sites. Its peak enrichment connects directly to the elevated immune neighbor fractions observed in MBRT. The top valley-enriched gene is **Clu** (clusterin, log2FC −0.25), a stress-responsive chaperone. Effect sizes are modest because: (a) n=1 limits power; (b) the stripe model's transition zones dilute signal; (c) many cells near zone boundaries experience intermediate dose. Note that p-values in this table are derived from Wilcoxon tests treating cells as independent observations; with n=1 tumor, cells share donor, batch, and microenvironment effects, making these p-values unreliable for biological inference. Effect sizes (log2FC, % expressed) are the appropriate measures.

## Rotation Null Validation

A critical control: the real stripe geometry (−25° angle, 1.05 mm spacing, 0.27 mm half-width) is rotated to 175 random angles on the same MBRT Block_21 tumor, and peak-valley deltas are recomputed at each angle. This preserves all spatial architecture (core vs edge, stromal infiltrate, TLS) while scrambling dose labels.

```{r rotation-null, results='asis'}
img(file.path(IF_DIR, "tumor_pv_rotation_null.png"),
    "Rotation null validation. Top: real MBRT delta distribution (red) exceeds rotation null (green) and SBRT null (grey). Bottom: per-gene scatter — 37 genes survive rotation null at empirical p<0.05.", width = "85%")
```

**37 of 670 tumor-expressed genes survive the rotation null** (empirical p < 0.05 AND |delta| > 2× overall null SD). Note: at p < 0.05 with 670 genes tested, ~33 genes would be expected by chance alone, so 37 survivors modestly exceeds the null expectation. The biological coherence of the surviving gene set — rather than the count — provides the stronger evidence:

- **Survivors enriched near peaks:** Tpt1, Uba52, B2m, H2-D1, H2-K1 (MHC class I), Ifitm2 (IFN response), Gpx1, Ppia, Itm2b, Fau, Tpm1, Gstp1/2
- **Rejected as architecture artifacts:** S100a6, S100a10, S100a4, Lgals1, Vim, Hsp90ab1, Naca (large MBRT delta but also large rotation-null SD — core-vs-edge tumor architecture confounding)

The surviving signature is dominated by **MHC class I induction, IFN response, damage/stress response, and translation machinery** — consistent with direct DNA damage at beam peaks. Future work with Mutter_02 replicates should increase the rotation null to ≥1000 rotations and apply FDR correction to strengthen this analysis.

## Cell-Type-Specific Peak-Valley DEGs

```{r pv-celltype-heatmap, results='asis'}
img(file.path(PLOT_DIR, "pv_celltype_heatmap.png"),
    "Top peak/valley DEGs by cell type. Color = log2FC (peak/valley).", width = "90%")
```

## Pathway Score Comparison

```{r pv-pathway, results='asis'}
pv_stats <- read_tsv(file.path(DATA_DIR, "pv_pathway_stats.tsv"), show_col_types = FALSE)
pv_stats %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(format = "html", caption = "Pathway scores: peak vs valley means and SDs") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

**Result: Pathway scores are indistinguishable between peak and valley zones at the bulk level.** The largest delta is 0.008 (DDR, Type I IFN) on a scale with SD ~1.0, yielding Cohen's d < 0.01 for all four pathways. STING scores are marginally *higher* in valleys (delta = −0.006) — while indistinguishable from noise, the direction is biologically plausible: cGAS-STING can be activated in bystander cells by cytoplasmic DNA from nearby damaged cells (via exosomes or gap junctions), which would preferentially occur in valley cells adjacent to peaks. This is a null result at current sample size — the 1000-gene panel's limited pathway gene coverage and/or the 4h timepoint may be insufficient to resolve pathway-level peak-valley differences. Testing with Mutter_02 replicates will determine whether this signal is real.

```{r pv-pathway-box, results='asis'}
img(file.path(PLOT_DIR, "pv_pathway_boxplots.png"),
    "Pathway scores by zone. Peak-valley differences are negligible (Cohen's d < 0.01 for all pathways). This is effectively a null result at the current sample size.", width = "70%")
```

```{r pv-pathway-celltype, results='asis'}
img(file.path(PLOT_DIR, "pv_pathway_by_celltype.png"),
    "Pathway scores stratified by cell type family and zone.", width = "90%")
```

## SBRT Comparison

```{r sbrt-heatmap, results='asis'}
img(file.path(PLOT_DIR, "pv_sbrt_heatmap.png"),
    "Peak/valley signature genes in SBRT_4h: mean expression of top peak-up and valley-up genes across MBRT peak, MBRT valley, and SBRT.", width = "60%")
```

```{r sbrt-pathway, results='asis'}
img(file.path(PLOT_DIR, "pv_sbrt_pathway_comparison.png"),
    "Pathway scores: MBRT peak vs MBRT valley vs SBRT_4h.", width = "80%")
```

## Composition: Peak vs Valley

```{r pv-composition, results='asis'}
img(file.path(PLOT_DIR, "pv_composition_diff.png"),
    "Cell type enrichment differences between peak and valley zones. Red = enriched in peaks, blue = enriched in valleys.", width = "75%")
```

## Spatial Neighborhoods: Peak vs Valley

```{r pv-nn, results='asis'}
img(file.path(PLOT_DIR, "pv_spatial_nn.png"),
    "Immune neighbor fractions by cell type: peak vs valley zones (k=20 nearest neighbors).", width = "90%")
```

---

# Cell-Type-Specific Dose Gradients

## Continuous Distance Analysis

Rather than binary peak/valley classification, cells can be analyzed on a **continuous distance-to-beam-center axis**. Spearman correlation of each gene's expression with this distance metric reveals gradients that the binary classification would miss.

```{r gradient-table}
gradients <- read_tsv(file.path(DATA_DIR, "gradient_genes_by_celltype.tsv"), show_col_types = FALSE)

gradients %>%
  select(cell_type, n_cells, peak_gene1, peak_cor1, valley_gene1, valley_cor1) %>%
  mutate(peak_cor1 = round(peak_cor1, 3), valley_cor1 = round(valley_cor1, 3)) %>%
  arrange(peak_cor1) %>%
  kable(format = "html",
        caption = "Top peak- and valley-enriched gene per cell type (Spearman rho with distance to beam center). Negative rho = peak-enriched.",
        col.names = c("Cell type", "n", "Peak gene", "rho", "Valley gene", "rho")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  column_spec(4, color = "red") %>%
  column_spec(6, color = "blue")
```

### B Cell p21 Gradient

B cells show a distinctive **bimodal "donut" profile** for Cdkn1a: expression peaks at ±0.2 mm from beam center, with a dip at the center itself. Several mechanisms could explain this pattern:

1. **Lethal dose gap (preferred):** Cells at beam center receive lethal dose and die before mounting a transcriptional response, while cells at the beam edge receive sub-lethal dose sufficient to activate p21.
2. **Survivorship bias:** Beam-center cells that survive lethal dose may be inherently radioresistant (e.g., quiescent cells), diluting the p21 signal. This can be tested by checking whether B cell density drops at beam center.
3. **Beam profile convolution:** The actual minibeam dose profile is not rectangular — it has a penumbra. The dip may partially reflect the Gaussian/trapezoidal beam shape convolved with cell density.
4. **Cell composition shift:** If radiosensitive B cell subsets are selectively killed at beam center, the remaining cells may be enriched for radioresistant subtypes that do not express p21.

These explanations are not mutually exclusive; all may contribute to the observed profile.

```{r bcell-gradient, results='asis'}
img(file.path(PLOT_DIR, "bcell_p21_signed.png"),
    "B cell Cdkn1a expression vs signed distance from beam center. Bimodal peaks at ±0.2 mm = donut profile.", width = "75%")
```

```{r bcell-bax, results='asis'}
img(file.path(PLOT_DIR, "bcell_bax_signed.png"),
    "B cell Bax (pro-apoptotic) gradient vs distance from beam center.", width = "75%")
```

### p21 Gradient Across Cell Types

```{r p21-all-celltypes, results='asis'}
img(file.path(PLOT_DIR, "p21_gradient_by_celltype.png"),
    "Cdkn1a expression vs distance from beam center, faceted by cell type. B cells and dendritic cells show the sharpest gradients.", width = "90%")
```

### Cancer Cell (4T1-filtered) Gradient

```{r cancer-gradient, results='asis'}
img(file.path(PLOT_DIR, "cancer_4t1_gradient.png"),
    "4T1-filtered cancer cell gene gradients. Clu (stress) enriched near peaks; Hmgb2 (proliferation) enriched in valleys.", width = "80%")
```

### Multi-Gene Spatial Score

Combining 15 peak-correlated genes (Hsd17b2, Clec7a, Itgb1, Tlr8, Tlr2, Tnfrsf10b, Cxcr4, Ccr7, Aif1, Hif1a, Atf3, Mki67, Cdkn1a, Gadd45b, Bax) into a composite Z-score improves the Spearman correlation with beam distance from rho = −0.04 (Cdkn1a alone in B cells sensu stricto) to rho = −0.15 (composite across all B lineage cells: B.cell + Memory.B + Plasmablast, n=1,580). Note: the composite includes Mki67 (a proliferation marker), which could contribute correlation strength through cell composition shifts (proliferating cells may concentrate near peaks where they are selectively killed) rather than direct damage response.

```{r multigene-map, results='asis'}
img(file.path(PLOT_DIR, "bcell_multigene_damage_map.png"),
    "Multi-gene composite damage score spatial map (B cells, KDE-smoothed). Red shading = beam peak zones.", width = "80%")
```

---

# Morphology Analysis

## Morphological Metrics

CosMx cell segmentation provides per-cell morphological metrics extracted from IF images: cell area (µm²), nuclear area (pixels), circularity, solidity, and aspect ratio. These provide orthogonal, non-transcriptomic readouts of radiation response.

```{r morph-summary}
morph <- read_tsv(file.path(DATA_DIR, "morphology_summary.tsv"), show_col_types = FALSE)

morph %>%
  filter(metric == "Area.um2") %>%
  mutate(across(where(is.numeric), ~round(., 1))) %>%
  kable(format = "html", caption = "Cell area summary statistics across treatment groups at 4h",
        col.names = c("Group", "Metric", "n", "Mean", "Median", "SD", "Q25", "Q75")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Peak vs Valley at 4h: No Morphological Difference

```{r morph-violins, results='asis'}
img(file.path(PLOT_DIR, "morphology_violins.png"),
    "Cell morphology violins across four groups at 4h. Peak and valley distributions are nearly identical.", width = "90%")
```

**4h is too early for bulk morphological differences between peak and valley.** Peak and valley cell areas are nearly identical (median 88.1 vs 86.9 µm²). Notably, MBRT peak nuclear area is slightly *larger* than valley (mean 3408 vs 3334 px, +2.2%), which is the opposite of what simple apoptosis-in-peaks would predict. This may reflect early G2/M arrest in sub-lethally irradiated peak cells (DNA content increase) or could be within noise. This is consistent with Harada et al. (J Radiat Res 2025), who found no morphological stripe pattern in minibeam-irradiated tissue until 12–16 hours post-exposure. However, apoptotic nuclear condensation in individual radiosensitive cells (lymphocytes) can begin within 2–4h at high doses; the bulk morphology may not resolve this signal given the heterogeneous cell composition.

## Kinetics: MBRT Nuclear Shrinkage vs SBRT Swelling

The most striking morphological finding emerges at **day 2**, not 4h:

```{r morph-kinetics, results='asis'}
img(file.path(PLOT_DIR, "morphology_nuc_kinetics.png"),
    "Nuclear area kinetics: MBRT (red) progressively shrinks; SBRT (blue) swells at day 2 then partially recovers. 23% divergence at day 2.", width = "80%")
```

```{r morph-area-kinetics, results='asis'}
img(file.path(PLOT_DIR, "morphology_kinetics.png"),
    "Cell area kinetics. SBRT shows day 2 swelling (105 vs 82 µm² control) consistent with G2/M arrest-driven hypertrophy.", width = "80%")
```

**Interpretation:** MBRT median nuclear area declines progressively across timepoints while SBRT nuclei enlarge at day 2 then partially recover. The ~23% MBRT-vs-SBRT nuclear area divergence at day 2 is consistent with — but does not prove — qualitatively different damage responses. Possible mechanisms include:

- **SBRT:** G2/M arrest → nuclear swelling (cells accumulate DNA content but cannot divide). However, SBRT day 2 swelling could also reflect inflammatory edema or general tissue swelling rather than pure cell-cycle-driven hypertrophy.
- **MBRT:** Several non-exclusive explanations: (a) apoptosis/pyknosis in peak cells → nuclear condensation; (b) **survivorship bias** — selective loss of large, lethally irradiated cells from peaks leaves smaller surviving cells, mimicking a shrinkage signal; (c) mitotic catastrophe (aberrant mitosis followed by fragmentation) may transiently inflate nuclear area before resolving.

**Caveat on Circularity metric:** The CosMx-reported Circularity values in `morphology_summary.tsv` show physically impossible values (mean 1.66, SD 258.9 for Control), indicating data quality issues with this specific metric. Nuclear Area and Cell Area metrics are used for kinetic analysis; Circularity and Solidity should be interpreted cautiously until the source of these anomalies is resolved.

```{r morph-celltype, results='asis'}
img(file.path(PLOT_DIR, "morphology_area_by_celltype.png"),
    "Cell area by cell type across groups at 4h. No cell-type-specific morphological differences emerge at this early timepoint.", width = "95%")
```

---

# p21 Kinetics and Spatial Biology

## p21 Kinetics Across Cell Types and Treatments

```{r p21-kinetics, results='asis'}
img(file.path(IF_DIR, "cdkn1a_kinetics_mean.png"),
    "Cdkn1a (p21) kinetics per cell class. Log-normalized mean expression across control, MBRT (1h–6d), and SBRT (4h–6d). MBRT peak/valley subsets at 4h shown as star and cross.", width = "90%")
```

**Key observations:**

- Myeloid cells (macrophages, monocytes/DC) show the largest MBRT-vs-SBRT and peak-vs-valley p21 separation
- Tumor cells show strong bulk p21 induction but minimal peak/valley difference — consistent with the heterogeneous "a" bucket mixing true tumor cells with mesenchymal/endothelial populations
- B cells show high p21 at 4h with clear peak > valley separation
- p21 expression returns near baseline by day 6 for all cell types — though this could reflect either p21 resolution or selective death of p21-expressing cells (survivorship bias)

## Spatial B Cell p21 Mapping

```{r bcell-spatial, results='asis'}
img(file.path(PLOT_DIR, "bcell_p21_fraction_smoothed.png"),
    "Gaussian-smoothed fraction of Cdkn1a+ B cells (KDE, bw=250µm). Red bands = beam peak zones. Signal is real but sparse — 178 p21+ cells across the tissue.", width = "80%")
```

---

# Signature Projection

## Peak and Valley Signatures

Cell-type-stratified gene signatures were derived from the 4h peak-valley DEG analysis and projected across all timepoints to track signature evolution. **Important caveat:** Moran's I values for signature spatial autocorrelation are low (MBRT_4h: 0.079, MBRT_day2: 0.188), indicating weak spatial patterning at the single-cell level. The signature projection is exploratory and should be validated with Mutter_02 replicates.

```{r sig-kinetics, results='asis'}
img(file.path(PLOT_DIR, "signature_kinetics_bulk.png"),
    "Bulk peak and valley signature scores across timepoints.", width = "80%")
```

The peak signature shows highest activation at day 2, consistent with the secondary transcriptional wave observed in pathway kinetics. However, the signature does not clearly distinguish peak from valley cells at any timepoint, suggesting either the signature genes are not specific enough or that peak-valley differences are too subtle for single-cell signature scoring to resolve.

```{r sig-celltype, results='asis'}
img(file.path(PLOT_DIR, "signature_kinetics_celltype.png"),
    "Cell-type-stratified signature scores across timepoints.", width = "90%")
```

```{r sig-sbrt, results='asis'}
img(file.path(PLOT_DIR, "signature_sbrt_comparison.png"),
    "Peak/valley signatures in SBRT: is the valley transcriptome similar to uniform-dose SBRT?", width = "80%")
```

Comparing signature scores in SBRT to MBRT peak and valley zones tests whether the valley transcriptome resembles uniform low-dose irradiation. If valleys show bystander effects, their signature should diverge from both SBRT and control.

```{r sig-spatial, results='asis'}
img(file.path(PLOT_DIR, "signature_spatial_maps.png"),
    "Spatial maps of peak and valley signature scores.", width = "90%")
```

---

# Discussion

## First-in-Field Spatial Transcriptomics of MBRT

A targeted literature search (Scopus + PubMed, April 2026; search terms: "spatial transcriptomics" AND ("microbeam" OR "minibeam" OR "GRID" OR "lattice" OR "spatially fractionated")) returned **zero published studies** applying spatial transcriptomics to any spatially-fractionated radiation therapy modality. This project establishes the first spatial-transcriptomic peak-valley characterization of MBRT biology. Note: preprints on bioRxiv/medRxiv were not systematically searched and could potentially contain unpublished work in this space.

## Key Biological Insights

### 1. The Peak-Valley Transcriptomic Contrast Is Real

The rotation null validation establishes that peak-valley gene expression differences are not artifacts of tumor spatial architecture. Thirty-seven genes survive the null at the tumor level. The surviving signature — MHC class I, IFN response, stress/chaperone genes — is biologically coherent with direct DNA damage at beam peaks.

### 2. Cell-Type-Specific Dose Gradients Partially Recapitulate Known Radiosensitivity

The continuous gradient analysis detects gradients in dendritic cells, pericytes, and B cells that are consistent with radiosensitive cell types responding to the beam dose gradient. However, the precise cell-type ranking is unstable at n=1 — the top gradient gene and correlation magnitude differ substantially between cell types, and even the "strongest" correlations (|rho| ≈ 0.08–0.15) explain <3% of variance. Notably, the two macrophage labels show comparable gradient magnitudes (rho = −0.08 for Macrophage, −0.12 for PerC.macrophage) rather than the flat profile predicted by their classical radioresistance; this may reflect tissue-resident macrophage populations with different biology from classical radioresistant macrophages, or indirect effects via microenvironment changes. Formal radiosensitivity ranking awaits replicate data.

### 3. The p21 Donut Profile in B Cells

The bimodal p21 expression profile in B cells — peaking at ±0.2 mm from beam center with a dip at the center — is a novel observation. While the preferred interpretation is a dose-response nonlinearity (cells at beam center die before mounting a transcriptional response, creating a "lethal dose gap"), alternative explanations including survivorship bias, beam penumbra effects, and cell composition shifts cannot be ruled out (see Section 5.4). Testing whether B cell density drops at beam center would help distinguish these hypotheses.

### 4. MBRT and SBRT Produce Qualitatively Different Morphological Trajectories

The nuclear area kinetics divergence (MBRT shrinks, SBRT swells) is consistent with qualitatively different damage responses, but the mechanism is ambiguous at n=1. MBRT shrinkage could reflect apoptosis/pyknosis, survivorship bias (selective loss of large cells), or both. SBRT swelling could reflect G2/M arrest-mediated hypertrophy, inflammatory edema, or tissue-level swelling. The morphological observation is real; the mechanistic interpretation is provisional.

### 5. MBRT Shows Greater Spatial Immune-Tumor Mixing

At 4h, MBRT shows higher immune neighbor fractions than SBRT across cell types. This early timepoint likely reflects redistribution of resident tumor-infiltrating immune cells rather than de novo recruitment, which requires chemokine-mediated signaling over longer timescales. The peak-enrichment of Ccl8 (MCP-2, a monocyte chemoattractant activated by ATM→NF-κB signaling) suggests a mechanistic link: DNA damage in peaks triggers chemokine secretion that may drive later myeloid recruitment, potentially underlying the sustained immune mixing seen at day 2 and day 6.

### 6. Valley Biology: Bystander Effects Remain Uncharacterized

A critical gap in this analysis is the absence of a **valley-vs-control comparison**. The valley signature (Hmgb2, Ube2c, Pttg1 enrichment) could represent either genuine bystander signaling in low-dose tissue or simply unperturbed proliferating tumor. Distinguishing these requires testing whether valley cells are transcriptomically different from completely unirradiated control cells. Additionally, the canonical bystander signaling mediators — gap junction communication (Cx43/Gja1), ROS/RNS (Nfe2l2/Nrf2 targets, Hmox1), and DAMP release (HMGB1, calreticulin) — were not analyzed, in part due to limited panel coverage. Mutter_02 replicates should prioritize this comparison.

### 7. Differential Death Kinetics at 4h

A critical consideration for interpreting the 4h timepoint: immune and tumor cells die on different timescales after irradiation. B cells and T cells undergo rapid interphase apoptosis (hours), meaning peak-zone immune cell depletion may already be underway at 4h. In contrast, 4T1 tumor cells primarily die by mitotic catastrophe (days), so tumor cell death has not yet manifested at 4h. This differential timing means the 4h snapshot captures immune damage in progress but tumor damage largely pending — the day 2 and day 6 timepoints are essential for observing tumor cell death and the resulting composition shifts.

### 8. Clinical Translation Implications

These findings, if confirmed with biological replicates, suggest several translational implications for MBRT optimization:

- **Beam spacing:** The p21 donut profile implies a bystander signaling range of ~0.2–0.4 mm from beam edges. If bystander effects in valleys contribute to immune priming, wider beam spacing may optimize valley volume for immune activation.
- **Immunotherapy combination:** The MHC class I upregulation in peaks (H2-K1, H2-D1, B2m) and elevated immune-tumor mixing suggest MBRT may enhance antigen presentation. This provides a rationale for combining MBRT with anti-PD1/CTLA4 immunotherapy.
- **Timing of immune intervention:** The DDR pathway suppression at 1–4h (likely ATM/ATR-mediated global transcriptional shutdown during early repair) followed by day 2 rebound suggests a therapeutic window for immune modulation beginning at 24–48h post-irradiation.

## Limitations

- **n=1 per condition:** All analyses are exploratory and descriptive. Effect sizes are reported; p-values from Wilcoxon tests treating cells as independent observations are unreliable for biological inference (cells share donor, batch, and microenvironment).
- **Single timepoint for peak/valley:** Only 4h has H2AX validation. Peak/valley classification at other timepoints relies on signature projection, not direct dose verification.
- **1000-gene panel:** The CosMx Mouse UCC panel has known coverage gaps for radiation-specific genes and bystander signaling mediators (Cx43/Gja1, Hmox1, calreticulin). Negative results should be interpreted cautiously.
- **"a" bucket heterogeneity:** The insitutype "a" label encompasses diverse cell types. True 4T1 cancer cells are a subset (~38K of ~84K "a" cells). Bulk DEGs are dominated by this heterogeneous population.
- **No per-cell H2AX:** The CosMx IF channels were used for segmentation markers, not H2AX. H2AX data comes from chromogenic IHC on an adjacent section, so per-cell H2AX quantification is not available.
- **No valley-vs-control comparison:** The bystander hypothesis (valleys exhibit non-trivial transcriptomic changes from paracrine signaling) has not been directly tested by comparing valley cells to unirradiated control cells.
- **FOV batch effects not formally assessed:** Peak and valley FOVs occupy different physical tissue locations. FOV-to-FOV technical variation in CosMx could confound peak-valley biological differences, though the rotation null partially addresses this.
- **Adjacent section registration:** H2AX IHC on an adjacent serial section introduces 50–100 µm displacement uncertainty, meaningful relative to the 0.27 mm beam half-width.
- **Morphology data quality:** CosMx Circularity values show physically impossible values (mean > 1.0), indicating data processing issues with shape descriptors. Area-based metrics are used for interpretation but should be validated independently.

## Next Steps

1. **Mutter_02 replication:** 4 slides × 3 conditions at day 2, with biological replicates. H2AX IHC on adjacent sections in process. This is the critical step for converting exploratory findings into testable results.
2. **Valley-vs-control comparison:** Test whether valley cells are transcriptomically distinct from unirradiated control cells (bystander hypothesis).
3. **Cell density analysis:** Test whether cell density is lower in peaks than valleys at 4h — a simple prediction of radiation-induced cell death that would validate the peak/valley classification from a completely independent axis.
4. **FOV batch effect quantification:** Formal assessment of FOV-to-FOV technical variation to rule out this confound on peak/valley DEGs.
5. **Spatial variance decomposition:** Quantify what fraction of gene expression variance is explained by cell type, peak/valley zone, spatial position, and FOV batch.
6. **Ligand-receptor analysis:** CellChat/NicheNet between peak and valley zones to characterize intercellular communication at the dose boundary.
7. **Rotation null enhancement:** Increase to ≥1000 rotations with FDR correction for the replicate dataset.
8. **Beam dosimetry:** Obtain prescribed dose, energy, and PVDR from treatment planning to anchor biological interpretation to quantitative dose.

---

# Session Information

```{r session-info}
sessionInfo()
```

<style>
body { max-width: 1200px; margin: auto; }
figure { margin: 1.5em auto; }
figcaption { text-align: center; }
.main-container { max-width: 1200px; }
h1 { border-bottom: 2px solid #2c3e50; padding-bottom: 0.3em; }
h2 { border-bottom: 1px solid #ddd; padding-bottom: 0.2em; }
table { font-size: 0.9em; }
</style>
